---
title: 'COLA Registered Report: The brain basis of inconsistent language lateralisation (POWER ANALYSIS)'
author: The COLA Consortium (Dorothy V. M. Bishop, David Carey, Margriet Groen, Jessica
  Hodgson, John Hudson, Emma Karlsson, Mairéad MacSweeney, Adam J. Parker, Nuala Simpson,
  Paul A. Thompson, Kate E. Watkins, Zoe V. J. Woodhead)
date: "04/03/2020"
output: html_document
bibliography: power.bib
---




```{r Rpackage_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(simsem)
library(lavaan)
library(semPlot)

```

## Power estimate 

There are two considerations for power in this registered report: 1. Power to fit individual models with sufficient fit inidices, whether one factor or two factor; 2. Power to detect mispecified model (one factor) vs the 'true' model (two factor). The first power consideration can be shown to be satisfied by our proposed sample size by following Monte Carlo simulation results presented by @Wolf_2013. This paper presents results from a range of simple CFA models, varying numbers of factors, numbers of measured variables, and magnitudes of factor loadings.


### Power estimate for the model comparison of proposed 'true' model vs proposed mispecified model.

We can calculate the power to detect the model mispecification using simulation and the distribution of fit indices. We use the `R` package `simsem` [@Jorgensen_2018] to calculate the power for the model comparison. 


```{r simsem_mispec,echo=FALSE,message=FALSE,warning=FALSE}
#https://github.com/simsem/simsem/wiki/Example-21:-Power-of-Rejecting-Misspecified-Models-with-Varying-Sample-Size

myN<-400
#--------------------------------------------------------------#
#Setup factor loadings matrix
loading.null <- matrix(0, 6, 2)
loading.null[1:3, 1] <- NA
loading.null[4:6, 2] <- NA

#Setup starting values for parameter loadings
LY.NULL <- bind(loading.null, 0.7)

#Setup correlation matrix between Latent variables
latent.cor.null <- matrix(NA, 2, 2)
diag(latent.cor.null) <- 1

#Setup starting values for latent variables correlation
RPS <- binds(latent.cor.null, 0.6)

#Setup error structure of the observed (manifest) variables.
error.cor.mis <- matrix("rnorm(1, 0, 0.1)", 6, 6)
diag(error.cor.mis) <- 1
RTE <- binds(diag(6), misspec = error.cor.mis)


#Fit null CFA model
CFA.Model.NULL <- model(LY = LY.NULL, RPS = RPS, RTE = RTE, modelType = "CFA")

#--------------------------------------------------------------#
#--------------------------------------------------------------#

#Alternative model setup - with misspecification

loading.alt <- matrix(0, 6, 1)
loading.alt[1:6, 1] <- NA

LY.ALT <- bind(loading.alt, 0.7)
RPS.ALT <- binds(diag(1))
CFA.Model.ALT <- model(LY = LY.ALT, RPS = RPS.ALT, RTE = RTE, modelType="CFA")

#--------------------------------------------------------------#
#--------------------------------------------------------------#

Output.NULL <- sim(NULL, n = 50:600, CFA.Model.NULL,silent=TRUE)
Output.ALT <- sim(NULL, n = 50:600, CFA.Model.ALT, generate = CFA.Model.NULL,silent=TRUE)

cutoff <- getCutoff(Output.NULL, alpha = 0.05, nVal = myN)
plotCutoff(Output.NULL, alpha = 0.05)

```

```{r simsem_mispec2,echo=TRUE,message=FALSE,warning=FALSE}

getPowerFit(Output.ALT, nullObject = Output.NULL, alpha = 0.05, nVal = myN)
getPowerFit(Output.ALT, cutoff = cutoff, nVal = myN, condCutoff = TRUE)
```

```{r simsem_mispec3,echo=TRUE,message=FALSE,warning=FALSE}

plotPowerFit(Output.ALT, Output.NULL, alpha = 0.05)
plotPowerFit(Output.ALT, Output.NULL, alpha = 0.05, logistic = FALSE)
```

```{r simsem_mispec4,echo=TRUE,message=FALSE,warning=FALSE}

cutoff2 <- c(RMSEA = 0.05, CFI = 0.95, TLI = 0.95, SRMR = 0.06)
getPowerFit(Output.ALT, cutoff = cutoff2, nVal = myN, condCutoff = FALSE)
```

```{r simsem_mispec5,echo=TRUE,message=FALSE,warning=FALSE}

plotPowerFit(Output.ALT, cutoff = cutoff2)
plotPowerFit(Output.ALT, cutoff = cutoff2, logistic = FALSE)
```

